---
title: "fivethirtyeight2016analysis"
author: "Talia Duffy, taliagd2"
date: "2024-11-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
df = read.csv("fivethirtyeight2016.csv")
head(df)
```
trump change is the difference between trump's actual performance and his 538 predicted vote share.
negative numbers means 538 underpredicted trump's performance.

```{r}
df$trump_change = df$trump_538 - df$trump_actual
df$clinton_change = df$clinton_538 - df$clinton_actual
head(df)
```

What is the percentage of states that underpredicted?

```{r}
sum(df$trump_change < 0) / nrow(df)
```

states sorted by how much they missed trump

```{r}
library(tidyverse)
df |>
  arrange(trump_change)
```

```{r}
df |>
  arrange(clinton_change)
```



How much did they miss trump in swing states?

Wisconsin, Michigan, Pennsylvania, NH, Nevada, North Carolina, Florida, Ohio, Arizona, Iowa, Georgia, Minnesota, Colorado, Maine

```{r}
swing_states = c("Wisconsin", "Michigan", "Pennsylvania", "New Hampshire", "Nevada", "North Carolina", "Florida", "Ohio", "Arizona", "Iowa", "Georgia", "Minnesota", "Colorado", "Maine statewide")

df_swing_states = df |>
  filter(state %in% swing_states) |>
  arrange(trump_change)

df_swing_states
```


```{r}
ggplot(data = df_swing_states) +
  aes(
    x = reorder(state, trump_change), 
    y = trump_change,
    fill = trump_change < 0
  ) +
  geom_bar(stat = "identity") +
  scale_fill_manual(labels = c("overpredicted", "underpredicted"), values = c("lightgreen", "firebrick"))+
  geom_hline(yintercept = 0, col = "black")+
  scale_y_continuous(breaks = seq(-5,1,0.5)) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme_bw() +
  theme(
    plot.title=element_text(face="bold"), 
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "inches"),
    axis.title.x = element_text(margin=margin(t=10)),
    axis.title.y = element_text(margin=margin(r=15))
  ) +
  ggtitle(
    "How Much Did 538 Miss Trump in Key 2016 Swing States?",
    subtitle = "Difference Between Predicted and Actual Vote Percentage"
  ) +
  xlab("State") +
  ylab("Percentage Point Difference") +
  labs(fill = "")
  
```
```{r}
usa = map_data("state")
usa = usa |>
  rename(state = region)
df$state = tolower(df$state)
df = df |>
  mutate(state = str_replace_all(string = state, pattern = " statewide", replacement = "")) |>
  mutate(state = str_replace_all(string = state, pattern = "d.c.", replacement = "district of columbia"))
```

```{r}
df_usa_merge = inner_join(df, usa)
```


```{r}
ggplot(data = df_usa_merge) + 
  aes(x = long, y = lat, group = group, fill=trump_change) +
  geom_polygon(color="black") +
  scale_fill_gradient2(
    name="Percentage Point Difference", 
    low = "firebrick", 
    mid = "white",
    high = "green3",
    limits = c(-9,9), 
    breaks=c(-9,-4.5,0,4.5, 9), 
    na.value = "grey50"
  ) +
  theme_void() +
  theme(
    plot.title=element_text(face="bold"), 
    plot.margin = unit(c(0.25, 0.25, 0.25, 0.25), "inches"),
    legend.position = "bottom"
  ) +
  ggtitle(
    "How Much Did 538 Miss Trump Nationwide?",
    subtitle = "Difference Between Predicted and Actual Vote Percentage"
  )
```


